---
title: "Chapter 13"
subtitle: "ЁЯз║ Grouped Summaries"
author: "by Lorraine Gaudio"
date:   "`r paste('Version', format(Sys.Date(), '%B %d, %Y'))`"
team: "Fall 2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

In Chapters 11 and 12, you learned how to subset and transform data with `dplyr` using functions such as `select()`, `filter()`, `mutate()`, and `case_when()`. In this chapter, you will extend that workflow by learning how to organize data into groups and then calculate summaries for those groups. You will use `count()` to make frequency tables, `summarize()` to compute statistics such as mean and standard deviation, and `group_by()` to tell R which subgroups to analyze separately. You will also practice computing multiple statistics at once and working with more than one grouping variable.

This chapter is about learning when and how to ask R for grouped summaries, and then explaining what those summaries mean. You will also practice curiosity as a learning skill. Curiosity helps you pause before running code, predict what the output might show, check whether the result matches your question, and decide when a side question is worth exploring or when it is better to stay focused on the chapter goal.

At the same time, you will continue building reproducible workflow habits in Quarto by using headings, rendering often, checking your output, and memoing what your code is doing and what you learned from it.

By the end of Chapter 13 you will be able to:

* ЁЯзо Use `count()` to create frequency tables for one categorical variable.

* ЁЯФв Use `count()` with more than one variable to count combinations of categories.

* ЁЯУК Use `summarize()` to calculate summary statistics such as mean, standard deviation, maximum, and row count.

* ЁЯСе Use `group_by()` to compute separate results for each group instead of one result for the whole dataset.

* ЁЯФН Explain how grouping changes the meaning of a summary result.

* ЁЯзй Combine `group_by()` and `summarize()` in a pipeline to calculate grouped summaries.

* тЮХ Compute multiple summary statistics in a single `summarize()` call.

* ЁЯП╖я╕П Distinguish between a grouping variable and a summary variable when analyzing data.

* тЪЦя╕П Choose when `count()` is the better tool and when `summarize(n = n())` is the better tool.

* тЩ╗я╕П Remove grouping after summarizing by using `.groups = "drop"` or `ungroup()`.

* ЁЯР▒ Practice curiosity by making predictions, checking output, asking useful follow-up questions, and deciding when to stay focused.

* ЁЯУЭ Continue using Quarto to organize code, output, checks, and written interpretation in a reproducible analysis document.

Keep these goals in mind as you move through each section.

# Setting Up Your Notes

In Chapter 11, you learned how to create a Quarto file, keep the YAML header, use Markdown headings, write memos outside code chunks, and control chunks with `#|` options. In Chapter 12, you added a table of contents, section numbering, and inline R code inside your sentences. In Chapter 13, you will keep practicing those Quarto habits while learning how to count categories and summarize data by groups.

## Create a Quarto File 

For those of you who **have GitHub** and a course project, navigate to your course folder and open your course project. This will open RStudio. Click "Pull" in the Git pane to get the latest version of the course materials. After completing work, save, stage, commit, and push. 

If you **don't have GitHub** or a course project, just open RStudio and set your working directory to your course folder.

Open a new Quarto document in your course folder and save it as: `chapter13_notes.qmd` (`File > New File > Quarto Document`).

## YAML Header

Copy and paste the YAML header from this file into your new Quarto document. Update the title, author, and date fields in the YAML header.

```{r echo=FALSE, results='asis'}
cat("```yaml
---
title: \"Chapter 13 Notes\"
subtitle: \"group_by() and summarize()\"
author: \"YOUR NAME\"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
---
```")
```

## Headings

Use the section headings in this file to create the structure of your notes. You can copy and paste the section headings into your new Quarto document.

```markdown

# Loading Packages

# Counts, Center, and Spread

# count()

# summarize()

# group_by()

# Multiple Statistics

# Multiple Variables

```

Work through the sections under each heading. ЁЯТб **Tip:** Render to HTML often to see the results.

## ЁЯУж Packages

Install once (if needed): `install.packages("dplyr")`; `install.packages("dslabs")`

ЁЯОп Load the packages 

```{r message=FALSE, warning=FALSE}
# тЪб Type this in a Quarto code chunk:
library(dslabs)  # Data science labs package
library(dplyr)   # Data manipulation package
```

# ЁЯУЦ  Counts, Center, and Spread

Before using `summarize()` and `count()`, it helps to review a few data words you have likely seen before. You do **not** need to calculate these by hand in this chapter. You do need to understand what they mean so you can choose the right function and explain your results.

ЁЯР▒ Curiosity as an intellectual virtue means choosing where to spend your attention. If these review terms feel rusty, pause here and use this section to refresh them. If you already practiced these ideas (e.g., in the Data Understanding Notebook activity), you can skip this review section and [jump to dplyr's `count()`](#dplyr-count). Return if you need a reminder.

## Data Terminology

A **numeric variable** stores numbers that can be measured or calculated, such as `mpg`, height, or test score.

A **categorical variable** stores labels or group names, such as state, disease, or cylinder count.

In this chapter, we often **group by a categorical variable** and then **summarize a numeric variable**.

## Count

A **count** tells you how many observations there are.

- For a whole dataset, count means the total number of rows.

- For grouped data, count means the number of rows in each group.

If you want to know how many cars have 4 cylinders, 6 cylinders, or 8 cylinders, you are asking for a **count**.

## Center

**Center** describes a typical value.

### Mean

The **mean** is the average.

It adds the values and divides by how many values there are.

If the mean `mpg` for cars with 4 cylinders is higher than the mean `mpg` for cars with 8 cylinders, then the 4-cylinder cars are more fuel efficient on average.

### Median

The **median** is the middle value when the data are sorted.

ЁЯкЮ Median is useful when extreme values might pull the mean too high or too low.

## Spread

**Spread** describes how much the values vary.

### Standard Deviation

The **standard deviation** tells you how far values tend to be from the mean.

- A **small** standard deviation means the values are packed closely together.

- A **large** standard deviation means the values are more spread out.

If one cylinder group has a much larger standard deviation for `mpg`, that group has more variability in fuel efficiency.

### Range

The **range** is the distance from the smallest value to the largest value.

ЁЯкЮ Range gives a quick sense of spread, but it only uses the minimum and maximum values.

---

ЁЯР▒ The goal is to know which summary tool matches the question. 

- use **count** when you want to know how many rows are in a dataset or group

- use **mean** when you want an average

- use **standard deviation** when you want to describe spread

- use a **categorical variable** to make groups

- use a **numeric variable** when calculating statistics like mean or standard deviation

We'll start with counting categories with `count()` because counts help you see what categories are present before you calculate more complex summaries.

# dplyr's count() {#dplyr-count}

You can use `count()` to quickly count rows for each category or combination of categories.

`count()` is especially useful when you want a **frequency table**.

`?count`

ЁЯУЬ The SYNTAX

`data %>% count(variable_name)`

Or, for more than one grouping variable:

`data %>% count(variable1, variable2, ...)`

## Count one variable

ЁЯОп Count how many cars have each cylinder value.

```{r}
# тЪб Type this in a Quarto code chunk:
mtcars %>%
  count(cyl)
```

ЁЯЧг The code above counts how many rows fall into each value of `cyl`. The result is a new table with one row per cylinder value and a column named `n` that stores the counts.

ЁЯР▒ This is the curiosity move here: before I compare or explain anything, I want to know what categories are in the data and how many rows belong to each one. This step helps me understand the data before I ask a bigger data question.

ЁЯТб **Test knit.** Does the table show one row for each cylinder value and a column named `n`?

### count() versus length()

`length()` counts how many values are in a vector.

`count()` counts rows in a data frame for each category or combination of categories.

### count() with sorting

Sometimes you want to see the biggest groups first.

ЁЯОп Count cars by cylinder and sort from largest count to smallest count.

```{r}
# тЪб Type this in a Quarto code chunk:
mtcars %>%
  count(cyl, sort = TRUE)
```

ЁЯЧг `sort = TRUE` arranges the result so the largest count appears first. The counts are the same as before; only the row order changed.

## Count multiple variables

You can also count combinations of categories. 

ЁЯР▒ This is a good time to indulge curiosity a little further. Counting one variable told us how many cars had each cylinder value. Counting two variables lets us ask a more specific question: how are cylinder values distributed across transmission types?

ЁЯОп Count how many cars fall into each combination of cylinder count and transmission type.

> ЁЯзР **Notice:** In `mtcars`, `am` is a coded transmission variable. A value of `0` means automatic, and a value of `1` means manual.

```{r}
# тЪб Type this in a Quarto code chunk:
mtcars %>%
  count(cyl, am)
```

ЁЯЧг `count(cyl, am)` creates one row for each unique combination of `cyl` and `am`, then counts how many rows are in each combination.

---

This is useful when one grouping variable is not enough. 

If our question is only тАЬHow many rows are in each category?тАЭ, `count()` is enough. But what if our question asks about averages, maximums, spread, or other statistics? We'll need `summarize()`.

# dplyr's summarize()

You can use `summarize()` to compute summary statistics like the mean, median, etc.

`?summarize` 

ЁЯУЬ  The SYNTAX

`data %>%` 
  `summarize(new_column_name = some_operation(column_name))`
  
ЁЯОп Find the overall mean mpg for all cars.

```{r}
# тЪб Type this in a Quarto code chunk:
mtcars %>% 
  summarize(mean_mpg = mean(mpg))
```

ЁЯЧг The code above creates a new column called `mean_mpg` that contains the mean of the `mpg` column from the `mtcars` dataset.

**We name the new column using `new_name = calculation`.**

ЁЯОп Find several summary statistics for mpg: mean, max, and count of cars.

```{r}
# тЪб Type this in a Quarto code chunk:
mtcars %>% 
  summarize(mean_mpg = mean(mpg),
            max_mpg = max(mpg),
            count = n())
```

> ЁЯзР **Notice:** Do not put pipes `%>%` inside `summarize()`. Inside the function, separate each summary calculation with a comma `,`.

ЁЯТб **Test knit.** Does the output show one row with three summary columns: `mean_mpg`, `max_mpg`, and `count`?

---

ЁЯдФ `summary()` is a base R function that gives a quick overall description of an object, such as basic statistics for each variable in a data frame. `summarize()` from `dplyr` is used to create your own summary results, usually after `group_by()`, such as the mean or count for each group.

## count() versus summarize()

You can often get the same result in two ways.

ЁЯОп Use group_by() and summarize() to produce the same counts by cylinder.

```{r}
# тЪб Type this in a Quarto code chunk:
mtcars %>%
  group_by(cyl) %>%
  summarize(n = n())
```

ЁЯЧг This gives the same idea as `count(cyl)`. The difference is that `count()` is shorter and faster to write when your main goal is just to count rows.

Use `count()` when you mainly want frequencies. Use `group_by()` with `summarize()` when you want to calculate other statistics too, such as mean and standard deviation.

# dplyr's group_by()

`group_by()` changes the behavior of summary functions.

ЁЯУЬ The SYNTAX

`data %>%` 
  `group_by(variable1, variable2, ...) %>%`
  `some_operation()`

ЁЯОп Find the mean mpg for each cylinder count.

```{r}
# тЪб Type this in a Quarto code chunk:
cyl_means <- mtcars %>%
  group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg))
```

ЁЯЧг `group_by(cyl)` splits the cars into groups based on cylinder count, and `summarize()` calculates `mean(mpg)` for each group. `mean_mpg` is the name of the new summary column.

> ЁЯзР **Notice:** We chain `group_by()` and `summarize()` together using the pipe operator `%>%`. First, `group_by(cyl)` tells R to work separately by cylinder value. Then, `summarize()` calculates the mean `mpg` for each group.

```{r}
# тЬЕ 
cyl_means  
```

ЁЯР▒ This is the main curiosity move in Chapter 13. We are no longer asking, тАЬWhat is the overall mean?тАЭ We're asking, тАЬDoes the mean change across groups?тАЭ Grouping helps us ask a sharper question without leaving the learning goal.

ЁЯТб **Test knit**. Does cyl_means print as a small table with one row per cylinder value?

ЁЯдФ Grouping cars by `cyl` is useful because cars with different numbers of cylinders often have different fuel efficiency, so one overall average mpg would hide those differences. By grouping first, you can compare the typical mpg for 4-, 6-, and 8-cylinder cars separately.

## Explicit vs. Default Naming

If you do not give the summary column a name, **`dplyr` uses the summary expression itself as the column name** by default. So `summarize(mean(mpg))` creates a column named `mean(mpg)` instead of a cleaner name like `mean_mpg`.

ЁЯОп Find the mean mpg for each cylinder count without naming the summary column.

```{r}
# тЪб Type this in a Quarto code chunk:
mtcars %>%
  group_by(cyl) %>%
  summarize(mean(mpg))      
```

ЁЯЧг The default column name is the code itself: `mean(mpg)`. Default names can be confusing when reading or sharing results, so naming summary columns makes tables clearer.

ЁЯР▒ Notice the problem and build the habit of naming summary columns clearly.

---

Now that you have seen how `group_by()` changes one summary calculation, you can ask for several summary statistics at the same time.

# Multiple Statistics {#multiple-stats}

ЁЯОп Create several summary columns in one `summarize()` call, separated by commas.

ЁЯР▒ This is useful curiosity: one statistic rarely tells the whole story. A mean tells me the center, a standard deviation tells me about spread, and `n()` tells me how many rows were used for each group.

*Use `?n` to learn about the `n()` function.*

```{r}
# тЪб Type this in a Quarto code chunk:
cyl_stats <- mtcars %>%
  group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg),
            sd_mpg   = sd(mpg),
            n        = n())
```


```{r}
# тЬЕ Verify
cyl_stats
```

ЁЯЧг `n()` counts how many rows are in each group created by `group_by()`. 

ЁЯдФ It is similar to `count()` because both tell you how many observations are in a group, but `n()` is used inside `summarize()` when you are also calculating other statistics.

> ЁЯзР **Notice:** `length()` counts the number of elements in an object. For a vector, that means the number of values. For a data frame, `length()` counts columns, not rows. That is why `n()` is the clearer tool here.

ЁЯТб **Test knit.** Does the output show one row for each cylinder value and three summary columns: `mean_mpg`, `sd_mpg`, and `n`?

---


# Multiple Variables {#multiple-vars}

`group_by()` with one column creates one group for each unique value in that column. `group_by()` with two columns creates one group for each unique combination of values.

ЁЯР▒ This is a good moment to make a prediction before running the code. Curiosity is pausing long enough to make a reasonable guess, then checking that guess against the output.

тЭУ How many groups will `group_by(cyl, am)` create?

ЁЯОп Predict how many groups will be created by `group_by(cyl, am)` before running the code.

```{r}
# тЬЕ Pre-check
# Recall that mtcars$cyl has three unique values:
sort(unique(mtcars$cyl))

# Recall that mtcars$am has two unique values:
sort(unique(mtcars$am))
```

ЁЯЧг Two variables with 3 and 2 unique values can create $3 \times 2 = 6$ possible groups. 

ЁЯОп Find the mean mpg for each combination of cylinder count and transmission type.

> ЁЯзР **Notice:** In `mtcars`, `am` is a coded transmission variable. A value of `0` means automatic, and a value of `1` means manual.

```{r message=FALSE, warning=FALSE}
# тЪб Type this in a Quarto code chunk:
mtcars %>%
  group_by(cyl, am) %>%
  summarize(avg_mpg = mean(mpg),
            count = n())
```

ЁЯЧг:

- Group 1: cyl = 4, am = 0 (4-cylinder automatic cars)

- Group 2: cyl = 4, am = 1 (4-cylinder manual cars)

- Group 3: cyl = 6, am = 0 (6-cylinder automatic cars)

- Group 4: cyl = 6, am = 1 (6-cylinder manual cars)

- Group 5: cyl = 8, am = 0 (8-cylinder automatic cars)

- Group 6: cyl = 8, am = 1 (8-cylinder manual cars)

ЁЯТб **Test knit.** Does the output show one row for each `cyl` and `am` combination? 

---

If you see a message about grouped output, keep going. The next section explains what that message means.

## ungroup()

When you summarize data that has more than one grouping variable, `dplyr` may keep part of the grouping structure in the result. That leftover grouping can affect later code, so it is important to check whether your summary table is still grouped.

ЁЯР▒ We are asking, тАЬWhat did R keep behind the scenes, and could that affect my next step?тАЭ

### A. Default behavior: keeps earlier groups

ЁЯОп Show that with two grouping variables, `summarize()` keeps the earlier group by default.

```{r}
# тЪб Type this in a Quarto code chunk:
# A) Default behavior: "drop_last" (keeps earlier groups)
cyl_gear_stats <- mtcars %>%
  group_by(cyl, gear) %>%
  summarize(mean_mpg = mean(mpg))   # no .groups set
```


```{r}
# тЬЕ  Verify
group_vars(cyl_gear_stats)  # what groups remain?
```

```{r}
# тЬЕ Verify
is_grouped_df(cyl_gear_stats) # expect TRUE
```

ЁЯЧг The default behavior is `.groups = "drop_last"`. That means `summarize()` drops the last grouping variable, `gear`, but keeps the earlier grouping variable, `cyl`.

YouтАЩll want to be explicit when you do not want any grouping left.

### B. Make it explicit inside `summarize()`

ЁЯОп Return a plain, ungrouped table by telling `summarize()` to drop all groups.

```{r}
# тЪб Type this in a Quarto code chunk:
# B) Make it explicit: drop *all* groups in the result
cyl_gear_stats_drop <- mtcars %>%
  group_by(cyl, gear) %>%
  summarize(mean_mpg = mean(mpg), .groups = "drop")
```

ЁЯЧг  The `.groups` argument starts with a period and `"drop"` must be written in quotation marks.

```{r}
# тЬЕ Verify
group_vars(cyl_gear_stats_drop)  # expect character(0)
```

```{r}
# тЬЕ Verify
is_grouped_df(cyl_gear_stats_drop) # expect FALSE
```

ЁЯЧг Using `.groups = "drop"` tells `summarize()` to remove all grouping after the summary operation, resulting in a plain data frame without any grouping structure.

ЁЯТб **Test knit.** Does `group_vars(cyl_gear_stats_drop)` return `character(0)` and does `is_grouped_df(cyl_gear_stats_drop)` return `FALSE`?

### C. Alternative: drop groups after summarizing

ЁЯОп Achieve the same ungrouped result by using `ungroup()` after `summarize()`

```{r}
# тЪб Type this in a Quarto code chunk:
# C) Alternative: drop groups after summarizing
cyl_gear_stats_ungroup <- mtcars %>%
  group_by(cyl, gear) %>%
  summarize(mean_mpg = mean(mpg)) %>%
  ungroup()
```


```{r}
# тЬЕ  Verify
group_vars(cyl_gear_stats_ungroup)  # expect character(0)
```

```{r}
# тЬЕ Verify
is_grouped_df(cyl_gear_stats_ungroup) # expect FALSE
```

ЁЯЧг Using `ungroup()` after `summarize()` removes all grouping from the resulting data frame. This gives the same ungrouped result as using `.groups = "drop"` inside `summarize()`.

ЁЯР▒ When we are done summarizing by groups, we want remove grouping for a plain table.

Pick one method and use it consistently:
- Use `.groups = "drop"` when you want to remove grouping inside `summarize()`.
- Use `ungroup()` when you want to remove grouping as a separate step after summarizing.

---

Speaking of summarizing, we've made it to the end of the chapter. Let's review what you learned.

# Summary

# Summary

In this chapter, you learned how to describe data by groups using `dplyr`. You reviewed the basic ideas of count, center, and spread so you could interpret grouped results, then used `count()` to make frequency tables, `summarize()` to compute statistics, and `group_by()` to split a dataset into subgroups before summarizing. You also practiced calculating multiple statistics in one call, using `n()` inside `summarize()` to count rows within groups, and working with combinations of grouping variables.

Most importantly, you learned that grouping changes what a summary means. Instead of one result for the whole dataset, you can produce separate results for each category or combination of categories. This is a core skill for later data analysis because analysts often need to compare subgroups, not just describe an entire dataset with one number.

ЁЯР▒ You also practiced curiosity as part of your workflow. In this chapter, curiosity meant pausing to ask what a summary is actually showing, predicting how many groups R might create, checking whether the output matched your expectation, and deciding when to investigate further versus when to stay focused on the current learning goal.

---

---

Table 1. Functions for grouped summaries

| Function           | Purpose                                                                                                                    |
| ------------------ | -------------------------------------------------------------------------------------------------------------------------- |
| `count()`          | Counts how many rows fall into each value or combination of values. Useful for frequency tables.                           |
| `group_by()`       | Creates groups based on one or more variables so later summary functions work separately for each group.                   |
| `summarize()`      | Creates a smaller summary table by calculating one or more summary statistics.                                             |
| `n()`              | Counts the number of rows in the current group when used inside `summarize()`.                                             |
| `mean()`           | Calculates the average of numeric values.                                                                                  |
| `sd()`             | Calculates the standard deviation, which describes spread around the mean.                                                 |
| `max()`            | Finds the largest value in a vector.                                                                                       |
| `unique()`         | Shows the distinct values in a vector. Useful for checking possible groups before summarizing.                             |
| `length()`         | Counts the number of elements in an object. For a vector, it counts values. For a data frame, it counts columns, not rows. |
| `.groups = "drop"` | Removes grouping inside `summarize()` so the result is returned as an ungrouped table.                                     |
| `ungroup()`        | Removes grouping after a grouped operation has already been created.                                                       |



# Chapter Terms

**Categorical Data**: Data whose values place observations into groups or categories. The categories may have no natural order (nominal) or may have an order with uneven gaps between categories (ordinal). In R, categorical data are often stored as factors.

**Center**: The middle or typical location of a dataset. Measures of center, such as the mean and median, summarize where values tend to cluster.

**Count**: The number of observations or values in a dataset, group, or category. In introductory statistics, count tells you how many cases are present.

**Frequency**: The number of times a value or category appears in a dataset.

**Mean**: The arithmetic average of a set of numbers, found by adding all values and dividing by the number of values.

**Median**: The middle value in an ordered dataset. If there is an even number of values, the median is the average of the two middle values.

**Numerical Data**: Data recorded as numbers, where the values have meaningful magnitudes and consistent intervals. Numerical data are also called *quantitative* data.

**Pipe**: An operator that passes the result of one step directly into the next step, letting you write code as a readable sequence of actions instead of nesting functions inside each other. In tidyverse code, the pipe is usually `%>%`, so `x %>% f()` means тАЬtake `x`, then apply `f()`тАЭ; in newer base R, a similar built-in pipe is `|>`.

**Range**: A simple measure of spread found by subtracting the smallest value from the largest value in a dataset.

**Spread**: How much the values in a dataset vary or scatter rather than cluster tightly together. Common measures of spread include range, interquartile range (IQR), variance, and standard deviation.

**Standard Deviation**: A measure of spread that describes how far values typically vary from the mean. A small standard deviation means values are clustered close to the mean, while a large standard deviation means they are more spread out.

# ЁЯУЭ Practice Space

In this Practice Space, you will use grouped summaries to describe data in a reproducible report. You will practice `count()` for frequency tables, `group_by()` for creating groups, `summarize()` for calculating statistics, and `n()` for counting rows inside grouped summaries. You will also keep practicing Quarto as a reproducible workspace for code, output, checks, and memo notes.

Create a new Quarto document and save it in your course folder as:

`chapter13_practice.qmd`

Before you begin:

1. Create a new Quarto document in RStudio.
2. Save it as `chapter13_practice.qmd`.
3. Edit the YAML so it includes the title `"Chapter 13 Practice"`, your name, the date, and HTML output.
4. Include the Chapter 13 Quarto settings for a table of contents, section numbering, and embedded resources.
5. Delete the default body text below the YAML.
6. Create headings for each task using `## Task 1`, `## Task 2`, and so on.
7. Load the `dplyr` and `dslabs` packages near the top of your document.
8. Render your document often so you can catch formatting problems and code errors early.

Complete each task in order. Fill in the blanks, run your code, check your output, and add short memo notes as you work.

In this Practice Space, you will see four memo prompts:

* ЁЯОп **Write the goal:** state the coding directions in your own words.
* ЁЯТм **Make a prediction:** predict what you expect before running the code.
* ЁЯЧг **Add Comments:** explain what the code or output shows.
* ЁЯкЮ **Reflect:** explain your thinking, what you checked, what changed, what confused you, or what you would try next.

Each task should include at least one memo response. You do not need to answer every prompt perfectly. The goal is to practice documenting your thinking while you work.

**Open-mindedness** means comparing reasonable methods and choosing the approach that makes the code easiest to read, check, and explain. If you encounter an error, document your **tenacity** by briefly noting what happened, what you tried, and how you fixed it or what still confuses you. Document how you troubleshot the error and what resource(s) you used to help you resolve the issue (**transparency**). Show **courage** by sharing your thinking process, even if it includes mistakes, confusion, or a question you are still working through. Show **curiosity** by noticing something worth understanding, asking a useful question, checking what R returns, and deciding whether to investigate further or stay focused on the current task.

You may quote AI output only when you clearly mark it as quoted text, either with quotation marks or a block quote using `>`. Then hand-type your own explanation of why you are including the quote. Type all code yourself.

Before you submit, clear your Environment and render the `.qmd` from top to bottom. Your file should render without errors and show your code, output, checks, and memo notes.

For this assignment, submit the rendered `chapter13_practice.html` file.

---

## Task 1

ЁЯМЯ **Single grouping**

Using the `stars` dataset, compute `mean_temperature` and `mean_magnitude` for each star `type`. Name the result `twinkle_twinkle`.

ЁЯР▒ This is a good place to practice focused curiosity. If you are curious about the star types, use `?stars` or `unique(stars$type)` to understand what groups exist. Then stay focused on the main task: checking whether `group_by()` created one summary row for each group.

```{r, eval=FALSE}
# тЬЕ Pre-check
data(stars)
?stars
colnames(stars)
unique(stars$type)
```

ЁЯЧг **Add Comments:** Which column is the grouping variable? Which columns are the numeric summary variables?

ЁЯОп **Write the goal:** Restate the goal of this task in your own words.

```{r, eval=FALSE}
# тЪб Code
twinkle_twinkle <- stars %>%
  _____(_____) %>%
  summarize(
    mean_temperature = ____(_____),
    mean_magnitude   = ____(_____),
    .groups = "drop"
  )
```

```{r, eval=FALSE}
# тЬЕ Check
twinkle_twinkle
is_grouped_df(twinkle_twinkle)
```

ЁЯЧг **Add Comments:** Does the output show one row for each star type? What does each summary column represent?

ЁЯкЮ **Reflect:** How did grouping by `type` change the meaning of the mean compared with calculating one mean for the whole dataset?

**ЁЯТб Test knit.**

---

## Task 2

ЁЯлТ **Single grouping + three statistics**

Pick one numeric fatty acid in the `olive` dataset, such as `palmitic`, `oleic`, or another fatty-acid variable. Group by `region`, then calculate:

* the mean of your chosen fatty acid,
* the standard deviation of your chosen fatty acid, and
* the number of rows in each region.

Store the result as `Olive_Garden`.

```{r, eval=FALSE}
# тЬЕ Pre-check
data(olive)
?olive
colnames(olive)
unique(olive$region)
```

ЁЯЧг **Add Comments:** Which column will you use as the grouping variable? Which fatty acid column will you summarize?

ЁЯОп **Write the goal:** Restate the goal of this task in your own words.


```{r, eval=FALSE}
# тЪб Code
fatty_acid <- "_______" # You Pick!

# Pick from operators: `=`, `<-`, `%>%`, `|`, `&`, `,` `(`, and `)`
Olive_Garden <- olive ___
  group_by(region) ___
  summarize(
    mean_value = mean(.data[[fatty_acid]], na.rm = TRUE)__
    sd_value   = sd(.data[[fatty_acid]], na.rm = TRUE)__
    n          = n()__
    .groups = "drop"
  __
```

> ЁЯзР **Notice:** `[[ ]]` pulls one column from a data frame. If you typed `olive[["oleic"]]`, R would use the `oleic` column. Here, `fatty_acid` stores the column name as text, so `.data[[fatty_acid]]` tells `dplyr` to use whichever column name you saved in `fatty_acid`.

```{r, eval=FALSE}
# тЬЕ Check
Olive_Garden
arrange(Olive_Garden, desc(mean_value))
is_grouped_df(Olive_Garden)
```

ЁЯЧг **Add Comments:** Which region has the highest mean value for the fatty acid you selected? How many rows are in each region?

ЁЯкЮ **Reflect:** Why is `region` the grouping variable and the fatty acid column the summary variable?

**ЁЯТб Test knit.**

---

## Task 3

ЁЯзо **Counting unique values**

Use `unique()` and `length()` to determine how many distinct values appear in a column.

For the `us_contagious_diseases` dataset from `dslabs`, determine the number of unique diseases in the column `disease`.

```{r, eval=FALSE}
# тЬЕ Pre-check
data("us_contagious_diseases")
?us_contagious_diseases
colnames(us_contagious_diseases)
```

```{r, eval=FALSE}
# тЬЕ Check unique diseases
unique(us_contagious_diseases$disease)

# тЬЕ Count unique diseases
_____(______(us_contagious_diseases$disease))
```

ЁЯЧг **Inline memo sentence**

Copy and paste the text below as normal text in your Quarto document, not inside a code. Replace the blank with the correct function to count unique diseases.

```{r, eval=FALSE}
ЁЯЧг There are `r _____(_____(us_contagious_diseases$disease))` unique diseases within the `disease` column of the `us_contagious_diseases` dataset. They are `r paste(unique(us_contagious_diseases$disease), collapse = ", ")`.
```

ЁЯкЮ **Reflect:** Why do you need both `unique()` and `length()` here? What does each function contribute?

Now determine the number of unique states in the column `state`.

```{r, eval=FALSE}
# тЬЕ Check unique states
unique(us_contagious_diseases$state)

# тЬЕ Count unique states
_____(_____(us_contagious_diseases$state))
```

ЁЯЧг **Inline memo sentence**

Write this as normal text in your Quarto document, not as code. Replace the blank with the correct function to count unique states.

```{r, eval=FALSE}
ЁЯЧг There are `r _____(_____(us_contagious_diseases$state))` unique states within the `state` column of the `us_contagious_diseases` dataset.
```

**ЁЯТб Test knit.**

---

## Task 4

ЁЯТм **Make and test a prediction**

In this task, you will predict how many groups will be created when you group by both `state` and `disease`, then test your prediction with code.

### Step 1

If there are 51 states in the dataset, and you assume every state has data for all 7 diseases, how many groups will `group_by(state, disease)` create?

A. $51 + 7 = 58$ groups

B. $51 \times 7 = 357$ groups

C. $\frac{51}{7} \approx 7$ groups

D. $51 - 7 = 44$ groups

E. $51^7$ groups

ЁЯТм **Make a prediction:** Choose one answer and explain why you chose it before running the code.

### Step 2

Test your prediction.

Use `group_by()` with both `state` and `disease`, then use `count()` to count the number of rows for each state-and-disease combination. Store the result as `categorical_counts`.

ЁЯОп **Write the goal:** Restate the goal of this task in your own words.

```{r, eval=FALSE}
# тЪб Code
categorical_counts <- us_contagious_diseases %>%
  _____(state, disease) %>%
  _____()
```

```{r, eval=FALSE}
# тЬЕ Check
categorical_counts
_____(categorical_counts)
```

ЁЯЧг **Add Comments:** Was your prediction correct? How many rows did `categorical_counts` contain?

ЁЯкЮ **Reflect:** Why is grouping useful when reporting the frequency of categorical data like `state` and `disease`?

ЁЯР▒ If the row count is not what you expected, pause and ask what assumption might have been wrong.

**ЁЯТб Test knit.**

---

## Task 5

ЁЯжа **National mean counts by disease**

From `us_contagious_diseases`, calculate the national mean of `count` for each `disease`. Then keep only the three diseases with the highest national_mean_count. Store the result as `Infectious_Burden`.

ЁЯТб **Tip:** Arrange the diseases by their national mean and keep the top three.

ЁЯР▒ тАЬWhich disease has the highest national mean count?тАЭ 

ЁЯОп **Write the goal:** Restate the goal of this task in your own words.

```{r, eval=FALSE}
# тЪб Code
______________ <- ____________ %>%
  group_by(_____) %>%
  summarize(
    ______________ = mean(_____, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(_____)) %>%
  slice_head(n = 3)
```

```{r, eval=FALSE}
# тЬЕ Check
Infectious_Burden
nrow(Infectious_Burden)
is_grouped_df(Infectious_Burden)
```

ЁЯЧг **Add Comments:** Which three diseases had the highest national mean counts? How do you know the output kept only three rows?

ЁЯкЮ **Reflect:** Why does `group_by(disease)` need to happen before calculating the mean of `count`?

**ЁЯТб Test knit.**

---

## Task 6

ЁЯУЭ **Final reflection**

Write a short paragraph explaining why `group_by()` needs to come before `summarize()` when you want separate results for each group.

Your paragraph should explain the difference between:

* summarizing the whole dataset,
* summarizing within groups,
* using `count()` for frequency tables, and
* using `summarize()` when you want custom statistics.

ЁЯкЮ **Reflect:** Include one sentence about curiosity in this chapter. What did curiosity help you notice, predict, check, or decide not to pursue?

## Task 7

ЁЯз╣ **Self-check checklist**

For the file submission, make sure your `.qmd` renders to HTML without errors and that your report shows the required objects, checks, and memo notes:

* `twinkle_twinkle` exists and includes one row for each star `type`.

* `twinkle_twinkle` includes the columns `mean_temperature` and `mean_magnitude`.

* `twinkle_twinkle` is ungrouped after summarizing.

* `Olive_Garden` exists and includes one row for each `region`.

* `Olive_Garden` includes the columns `mean_value`, `sd_value`, and `n`.

* `Olive_Garden` is ungrouped after summarizing.

* Your report correctly identifies the number of unique diseases in `us_contagious_diseases$disease`.

* Your report correctly identifies the number of unique states in `us_contagious_diseases$state`.

* Your inline R code renders correctly in the memo sentences for unique diseases and unique states.

* `categorical_counts` exists and shows counts for each `state` and `disease` combination.

* Your report includes a prediction about how many `state` and `disease` groups you expected, followed by a check of the actual result.

* `Infectious_Burden` exists and contains only the three diseases with the highest national mean counts.

* `Infectious_Burden` includes the column `national_mean_count`.

* Your report includes short memo notes that show your goals, comments, checks, predictions, and reflections.

* Your final reflection explains why `group_by()` needs to come before `summarize()` when you want separate results for each group.

* Your report includes at least one place where you documented curiosity by noticing something interesting, asking a useful question, checking what R returned, or deciding to stay focused on the current task.

* All code in your Quarto document runs top-to-bottom without errors.

ЁЯОп Check your `.qmd` and rendered HTML for submission quality.

**Step 1: Sweep your Environment.**
Click the ЁЯз╣ **broom icon** in the Environment pane to clear objects.

**Step 2: Run everything top-to-bottom.**
Use **Run тЖТ Run All** to execute every chunk in order.

**Step 3: Render to HTML.**
Click **Render** and confirm that the HTML file is created without errors.

**Step 4: Verify the required objects and checks.**
Use the checklist above to confirm that each object exists and contains the expected rows, columns, summaries, counts, or grouped results.

**Step 5: Read your memo notes.**
Make sure your comments and reflections explain what your code did, what you checked, what you predicted, and what you learned. Verify that any inline R code worked in your memos. Look for at least one place where you documented curiosity by noticing something worth understanding, asking how something worked, checking the output, or deciding to stay focused.

**Step 6: Save and submit.**
After your Quarto document renders without errors, make sure your latest edits are saved. Submit the rendered `chapter13_practice.html` file as directed.


