---
title: "Missing Backtick Example Report"
output:
  html_document:
    df_print: paged
---

## Task 1: Missing Values

🎯 **Goal:** Create a vector with missing quiz scores, find where the missing values are located, and calculate the mean after removing missing values.

💬 **Prediction:** I think R will return `NA` if I try to calculate the mean before telling it how to handle the missing value.

```{r}
# ⚡ Code
quiz_scores <- c(84, 91, NA, 76, 88)

# ✅ Check whether the vector contains missing values
anyNA(quiz_scores)

# ✅ Find the position of the missing value
which(is.na(quiz_scores))

# ✅ Calculate the mean after removing missing values
mean(quiz_scores, na.rm = TRUE)

``

✍ **Memo:** I used `anyNA()` to check whether there was a missing value. Then I used `is.na()` inside `which()` to locate the missing value. I had to add `na.rm = TRUE` because `mean()` returns `NA` when the vector contains a missing value.

## Task 2: Bracket Indexing

🎯 **Goal:** Use bracket indexing to select specific quiz scores.

💬 **Prediction:** I think `quiz_scores[2]` will return the second value in the vector.

```{r}
# ⚡ Code
quiz_scores[2]

# ✅ Check selected values
quiz_scores[c(1, 4)]

# ✅ Check all scores except the missing value
quiz_scores[-3]
```

🗣 **Comment:** Bracket indexing lets me select values by position. Negative indexing removes positions from the result.

## Task 3: Summarizing Missing Values

🎯 **Goal:** Count how many missing values are in the vector and calculate the proportion of values that are missing.

```{r}
# ⚡ Code
sum(is.na(quiz_scores))
mean(is.na(quiz_scores))
```

🗣 **Comment:** `sum(is.na(quiz_scores))` counts how many values are missing. `mean(is.na(quiz_scores))` gives the proportion of values that are missing because `TRUE` is treated like 1 and `FALSE` is treated like 0.

## Troubleshooting Memo

❓ **Question:** Why did this report fail to knit?

👀 **Resource Cited:** I checked the Render tab error message and looked near the line number where the error appeared.

✍ **Memo:** The first code chunk starts with an opening code fence, but it does not have a matching closing code fence before the memo and Task 2 heading. R tries to read regular Markdown text as if it were R code. The fix is to add three backticks after the final line of code in Task 1, before the memo text.